---
title: "Classification using decision trees - Risk of Cancer in USA"
author: "Ambareesh Jonnavittula"
date: "29/12/2021"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)
```


## Contents

- Step 1 - Problem Definition
- Step 2 - Environmental Setup
- Step 3 - Data Acquisition
- Step 4 - Data Preparation 
- Step 5 - Exploratory Data Analysis
- Step 6 - Data Preprocessing 
- Step 7 - Classification - Decision tree
- Step 8 - Model Evaluation

## Step 1 - Problem Definition

Over the years, there have been various carcinogenic factors which have caused Cancer among the people across all the states of USA according to NATA.

The objective of this task is to be able to prepare a classification model to utilize the predictor variables to identify and/or predict which factors potentially pose a range of risk of cancer among the people of USA.

_Note - In addition, this piece of code is reproducible, hence we could scale it to any state of the USA within the dataset_

## Step 2 - Environmental Setup

- dplyr : Data manipulations
- tidyverse : Data science tasks
- readxl : to Import the .xlsx file
- skimr : Statistical summary
- corrplot : Correlation matrix
- rpart : Classification - Decision Trees 
- rattle - Decision tree visualization
- vioplot : Violin plots
- ggplot2 : Plotting graphs
- RColorBrewer : Colour palette

```{r include=FALSE}
# 2.1 - Install the required packages/libraries from R
# install.packages('dplyr') 
# install.packages('tidyverse')
# install.packages('ggplot2')
# install.packages('rpart')
# install.packages('rpart.plot')
# install.packages('rattle')
# install.packages('RColorBrewer')
# install.packages('skimr')
# install.packages("xlsx")
# install.packages("vioplot")
# install.packages("corrplot")
# install.packages("Hmisc")
# install.packages("ROCR")

# Libraries to current R session
library(dplyr) 
library(tidyverse)
library(ggplot2)
library(rpart)
library(rpart.plot)
library(rattle)
library(RColorBrewer)
library(skimr)
library("readxl")
library(vioplot)
library(Hmisc)
library(corrplot)

# 2.3 - Setup Working directory
fpath = "D:/University/ASDM/Developments/Task 1 - Classification"
setwd(fpath)   # Set the WD
getwd()        # Validate the new WD
```

## Step 3 - Data Acquisition

- Import data from the file path using read_excel() function
- Rename columns for simplification
```{r echo=TRUE}
## Importing / Reading the data into "df_cea_raw" data frame 
df_cea_raw <- read_excel("ARM Dataset.xlsx", sheet = 1) 

# Inspect the raw data
head(df_cea_raw)
tail(df_cea_raw)
names(df_cea_raw)
summary(df_cea_raw)
str(df_cea_raw)
dim(df_cea_raw)

# Renaming the columns as part of simplification
names(df_cea_raw)[names(df_cea_raw) == 'State'] <-  'state'
names(df_cea_raw)[names(df_cea_raw) == 'County'] <-  'county'
names(df_cea_raw)[names(df_cea_raw) == 'Pollutant.Name'] <-  'pollutant'
names(df_cea_raw)[names(df_cea_raw) == 'Point..includes.railyards..Cancer.Risk..per.million.'] <-  'railyards_crpm'
names(df_cea_raw)[names(df_cea_raw) == 'Airport.Cancer.Risk..per.million.'] <-  'airport_crpm'
names(df_cea_raw)[names(df_cea_raw) == 'OR.Lightduty..includes.refueling..Cancer.Risk..per.million.'] <-  'lightduty_crpm'
names(df_cea_raw)[names(df_cea_raw) == 'OR.Heavyduty.Cancer.Risk..per.million.'] <-  'heavyduty_crpm'
names(df_cea_raw)[names(df_cea_raw) == 'NR..no.airports..CMV..locomotives..Cancer.Risk..per.million.'] <-  'cmv_loco_crpm'
names(df_cea_raw)[names(df_cea_raw) == 'NP.10m.ReleaseHeight.Cancer.Risk..per.million.'] <-  'np_10m_releaseheight_crpm'
names(df_cea_raw)[names(df_cea_raw) == 'NP.Low.ReleaseHeight.Cancer.Risk..per.million.'] <-  'np_low_releaseheight_crpm'
names(df_cea_raw)[names(df_cea_raw) == 'ResidentialWoodCombustion..RWC..Cancer.Risk..per.million.'] <-  'rwc_crpm'
names(df_cea_raw)[names(df_cea_raw) == 'NR.CommercialMarineVessel..CMV..Cancer.Risk..per.million.'] <-  'cmv_crpm'
names(df_cea_raw)[names(df_cea_raw) == 'Biogenics.Cancer.Risk..per.million.'] <-  'biogenics_crpm'
names(df_cea_raw)[names(df_cea_raw) == 'Fires..ag..prescribed..and.wild..Cancer.Risk..per.million.'] <-  'fires_crpm'
names(df_cea_raw)[names(df_cea_raw) == 'Secondary.Cancer.Risk..per.million.'] <-  'secondary_crpm'
names(df_cea_raw)[names(df_cea_raw) == 'Total.Cancer.Risk..per.million.'] <-  'total_crpm'
```

## Step 4 - Data Preparation

1. Subset to consider only required columns & drop the raw dataframe
2. Inspect the subset
3. Create a function to Flatten the correlation matrix
4. Classify the level of threat into L0, L1, L2, & L3
```{r echo=TRUE}
# 4.1 Subset to consider only required columns
df_cea <- select(filter(df_cea_raw)
                 ,c(state, county, pollutant,
                    railyards_crpm,
                    airport_crpm,
                    lightduty_crpm,
                    heavyduty_crpm,
                    cmv_loco_crpm,
                    np_10m_releaseheight_crpm,
                    np_low_releaseheight_crpm,
                    rwc_crpm,
                    cmv_crpm,
                    biogenics_crpm,
                    fires_crpm,
                    secondary_crpm,
                    total_crpm  ))

# 4.3 To replace missing values with 0
df_cea[is.null(df_cea)] = 0

# 4.2 Drop the raw dataframe
remove(df_cea_raw)

```

3. Inspect the subset 
```{r echo=TRUE}
# 4.3 Inspect the subset
# Describing the data subset 
head(df_cea)
tail(df_cea)
names(df_cea)
str(df_cea)
summary(df_cea)

# Check the number of Rows & Columns in the data subset
nrow(df_cea) 
ncol(df_cea) 
dim(df_cea)   

skim(df_cea)

```

3. Function to Flatten the correlation matrix
```{r echo=TRUE}
# c is the corr. coeff. matrix
# p is the corr. p-values matrix
flat_cm <- function(c, p) {
  ut <- upper.tri(c)
  data.frame(
    row = rownames(c)[row(c)[ut]],
    column = rownames(c)[col(c)[ut]],
    cor  =(c)[ut],
    p = p[ut]
  )
}
```

4. Classify the level of threat
```{r echo=TRUE}
summary(df_cea$total_crpm)

df_cea$level_of_threat <- as.factor(ifelse(df_cea$total_crpm <= 0.5, 'L0',
                                ifelse(df_cea$total_crpm <= 5, 'L1',
                                       ifelse(df_cea$total_crpm <= 10, 'L2', 'L3'))))

# df_cea$level_of_threat <- cut(df_cea$total_crpm, 4, labels = c('L0', 'L1', 'L2', 'L3'))

# Check frequency
as.data.frame(table(df_cea$level_of_threat)) 
```

## Step 5 - Exploratory Data Analysis

1. Correlation plots with Pearson correlation coefficients
```{r echo=TRUE}
# 5.1 - Histogram - Total CRPM
hist(df_cea$total_crpm, xlab="Total Cancer Risk per Million", main = "Distribution of values - Total Cancer Risk per million")

# 5.2 - Box plot - Total Cancer risk by State
ggplot(df_cea, aes(x=state, y=total_crpm, fill=state)) +
  geom_boxplot()+
  labs(title="Box plot - Total Cancer risk per million by State",x="State", y = "Cancer Risk")+   
  theme(legend.position = "none") + 
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))+
  theme(plot.title = element_text(hjust = 0.5))

# 5.3 - Violin Plot - Total CRPM by Pollutant
ggplot(df_cea, aes(x=pollutant, y=total_crpm)) +
  geom_violin(trim=FALSE, fill='#A4A4A4', color="darkred")+
  geom_boxplot(width=0.1, outlier.shape = NA) +  theme_minimal()

# 5.4 Correlation Matrix for continuous variables
df_cea_cv <- df_cea[, c(4:16)]
df_cea_cor <- cor(df_cea_cv)
round(df_cea_cor,2)

# 5.5 Apply Flatten function here
rs<-rcorr(as.matrix(df_cea[, c(4:16)]))
flat_cm(rs$r, rs$P)

# 5.6 - Correlation Matrix Analysis using Heatmap
col <- colorRampPalette(c("grey", "white", "red"))(20)
heatmap(x = df_cea_cor, col = col, symm = TRUE)

# 5.7 - Correlogram
corrplot(df_cea_cor, type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45)

# 5.8 - Level of Threat
ggplot(df_cea, aes(x=total_crpm, fill=level_of_threat, color=level_of_threat)) +
  geom_histogram( alpha=0.5, position="identity")+
  scale_color_brewer(palette="Dark2")+
  scale_fill_brewer(palette="Dark2")+
  labs(title="Level of Threat",x="Total Cancer risk per million", y = "Frequency")+ 
  theme(plot.title = element_text(hjust = 0.5))
```

## Step 6 - Data Preprocessing
```{r echo=TRUE}
# Subset to only predictor and response variables.
df_cea_model <- df_cea[, 4:17]
names(df_cea_model)

# Write the .csv file for SAS task
write.csv(df_cea_model,"D:/University/ASDM/Developments/Task 1 - Classification/Cancer_Risk_USA_SAS.csv", row.names = TRUE)

# Random number generator
set.seed(28) 

# Use Sampling function to split the training vs. test datasets.
cea_split <- sample(2, nrow(df_cea_model),replace=TRUE, prob=c(0.8,0.2))
head(cea_split)

# Create train & test datasets
train_cea <- df_cea[cea_split==1,] 
test_cea <- df_cea[cea_split==2,] 

# Inspect the datasets & split
dim(train_cea)
dim(test_cea)
as.data.frame(table(train_cea$level_of_threat))
as.data.frame(table(test_cea$level_of_threat))

# Function to be used to predict the class and calculate accuracy
model_predict <- function(df, tree, v_title ) {
  
  # 1. Plot the tree
  fancyRpartPlot(tree, main = v_title)
  
  # 2. Cancer Risk Prediction
  risk_prediction <- predict(tree, df, type = 'class')
  
  # 3. Confusion matrix 
  cmat <- table(df$level_of_threat, risk_prediction)
  print('Confusion Matrix is created')

  # 4. Accuracy % & Error %
  acc <- sum(diag(cmat)) * 100 / sum(cmat)
  print(paste('Accuracy :', round(acc,2), '%'))
  print(paste('Error :', round(100 - acc,2), '%'))
  
  # 5. Precision, Recall & F1 Score
  v_diag <- diag(cmat) # number of correctly classified instances per class
  rowsums <- apply(cmat, 1, sum) # number of instances per class
  colsums <- apply(cmat, 2, sum) # number of predictions per class
  v_precision <- v_diag / colsums
  v_recall <- v_diag / rowsums
  v_f1 <- 2 * v_precision * v_recall / (v_precision + v_recall)
  df_prf1 <- data.frame(round(v_precision*100,2), round(v_recall*100,2), round(v_f1*100,2))
  colnames(df_prf1) <- c('Precision (%)', 'Recall (%)', 'F1 Score (%)')
  df_prf1
}

```

## Step 7 - Decision Trees Classifier - Train the model

1. Train the model using rpart()
2. Plot the tree using fancyRpartPlot()
3. Predict the model
4. Create a confusion matrix
5. Measure Accuracy Rate % & Error Rate %

```{r echo=TRUE}
# By default, rpart() function uses the Gini impurity measure to split the note. The higher the Gini coefficient, the more different instances within the node.

# 1. Understand the variables 
tree1 <- rpart(level_of_threat ~ np_low_releaseheight_crpm, data=train_cea, method="class")
fancyRpartPlot(tree1, main = "np_low_releaseheight_crpm")

tree2 <- rpart(level_of_threat ~ cmv_loco_crpm, data=train_cea, method="class")
fancyRpartPlot(tree2, main = "cmv_loco_crpm")

tree3 <- rpart(level_of_threat ~ lightduty_crpm, data=train_cea, method="class")
fancyRpartPlot(tree3, main = "lightduty_crpm")

tree4 <- rpart(level_of_threat ~ heavyduty_crpm, data=train_cea, method="class")
fancyRpartPlot(tree4, main = "heavyduty_crpm")

# 2. Train the model
rpart_tree <- rpart(level_of_threat ~ railyards_crpm+
                                      airport_crpm +
                                      rwc_crpm +
                                      cmv_crpm +
                                      biogenics_crpm +
                                      fires_crpm +
                                      secondary_crpm +
                                      np_10m_releaseheight_crpm + 
                                      np_low_releaseheight_crpm + 
                                      cmv_loco_crpm + 
                                      lightduty_crpm + 
                                      heavyduty_crpm,
                        data=train_cea, method="class")

# 3. Evaluate the trained model
model_predict(train_cea, rpart_tree, 'Training : Decision tree to identify the risk of Cancer')

# 4. Accuracy Tuning - (not needed if > 78~80%) - to identify more nodes
ctrl <- rpart.control(minsplit = 4,
    minbucket = round(5 / 3),
    maxdepth = 4,
    cp = 0)
tune_fit_train <- rpart(level_of_threat ~ railyards_crpm+
                                      airport_crpm +
                                      rwc_crpm +
                                      cmv_crpm +
                                      biogenics_crpm +
                                      fires_crpm +
                                      secondary_crpm +
                                      np_10m_releaseheight_crpm + 
                                      np_low_releaseheight_crpm + 
                                      cmv_loco_crpm + 
                                      lightduty_crpm + 
                                      heavyduty_crpm,
                        data=train_cea, method="class", control = ctrl)

model_predict(train_cea, tune_fit_train, 'Trained & Tuned : Decision tree to identify the risk of Cancer')


```


## Step 8 - Decision Trees Classifier - Test the model

1. Test the model using rpart()
2. Compare the results with Training dataset.

```{r echo=TRUE}
# 1. Test the model
tune_fit_test <- rpart(level_of_threat ~ railyards_crpm+
                                      airport_crpm +
                                      rwc_crpm +
                                      cmv_crpm +
                                      biogenics_crpm +
                                      fires_crpm +
                                      secondary_crpm +
                                      np_10m_releaseheight_crpm + 
                                      np_low_releaseheight_crpm + 
                                      cmv_loco_crpm + 
                                      lightduty_crpm + 
                                      heavyduty_crpm,
                        data=test_cea, method="class", control = ctrl)

# 2. Evaluate the tested model
model_predict(test_cea, tune_fit_test, 'Tested : Decision tree to identify the risk of Cancer')

```












