---
title: "Association Rules Mining : Concomitant Pollutants in California, USA (NATA)"
author: "Ambareesh Jonnavittula"
date: "28/12/2021"
output: slidy_presentation
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)
```

## Contents

- Step 1 - Problem Definition
- Step 2 - Environmental Setup
- Step 3 - Data Acquisition
- Step 4 - Data Preparation
- Step 5 - Exploratory Data Analysis
- Step 6 - Data Preprocessing
- Step 7 - Association rules mining
- Step 8 - Model Evaluation

## Step 1 - Problem Definition

Over the years, there have been various toxic pollutants which have caused Cancer among the people across the counties of California, USA.  

The objective of this task is to be able to identify if there is high risk of multiple carcinogens in association with each other in each of the tract in California, USA. 

_Note - In addition, this piece of code is reproducible, hence we could scale it to any state of the USA within the dataset._

## Step 2 - Environmental Setup

- dplyr : Data manipulations
- tidyverse : Data science tasks
- readxl : to Import the .xlsx file
- skimr : Statistical summary
- corrplot : Correlation matrix
- arulesViz : Assoc Rules - Visualization
- arules : Association Rules
- ggplot2 : Plotting graphs
- RColorBrewer : Colour palette

```{r include=FALSE}
# 2.1 - Install the required packages/libraries from R
# install.packages('dplyr')
# install.packages('tidyverse')
# install.packages('skimr')
# install.packages('arulesViz') 
# install.packages('arules') 
# install.packages("ggplot2") 
# install.packages("corrplot") 
# install.packages('stringr')

# 2.2 - Load the libraries
library(dplyr) 
library(tidyverse)
library(stringr)
library(ggplot2)
library(RColorBrewer)
library(skimr)
library(corrplot)
library(readxl)  # To read .xlsx files
library(arulesViz) # activate “arules” package
library(arules) # activate “arules” package

# 2.3 - Setup Working directory
fpath = "D:/University/ASDM/Developments/Task 2 - Association Rules Mining"
setwd(fpath)   # Set the WD
getwd()        # Validate the new WD
```

## Step 3 - Data Acquisition

- Import data from the file path using read_excel() function
- Rename columns for simplification
```{r echo=TRUE}
## Importing / Reading the data into "df_cea" data frame 
df_cea_raw <- read_excel("ARM Dataset.xlsx", sheet = 1) 

# Inspect the raw data
head(df_cea_raw)
tail(df_cea_raw)
names(df_cea_raw)
summary(df_cea_raw)
str(df_cea_raw)
dim(df_cea_raw)

# Renaming the columns as part of simplification
names(df_cea_raw)[names(df_cea_raw) == 'State'] <-  'state'
names(df_cea_raw)[names(df_cea_raw) == 'County'] <-  'county'
names(df_cea_raw)[names(df_cea_raw) == 'Tract'] <-  'tract'
names(df_cea_raw)[names(df_cea_raw) == 'Pollutant.Name'] <-  'pollutant'
names(df_cea_raw)[names(df_cea_raw) == 'Total.Conc'] <-  'total_conc'
```

## Step 4 - Data Preparation

1. Subset to a single state
2. Aggregate cancer risk data per category using averages per county
```{r echo=TRUE}
# 4.1 Subset to state CA
df_cea <- select(filter(df_cea_raw, 
                        state == "CA" &
                        tract != 0 & pollutant != '' )
                 ,c(tract, pollutant, total_conc))

# 4.2 Aggregate data using averages
df_cea_agg <- df_cea %>%
              group_by(tract, pollutant) %>%
              summarise_at(vars(c(1)), list(avg = mean))

# 4.3 String manipulations within Pollutant
df_cea_agg$pollutant <- str_to_title(df_cea_agg$pollutant)

# 4.4 Pivot pollutants and total conc columns
df_cea_piv <- df_cea_agg %>% 
  pivot_wider(names_from = pollutant, 
                     values_from = avg)

# 4.5 Convert tract into a categorical variable
df_cea_piv$tract <- as.factor(df_cea_piv$tract)
 
# 4.5 Drop the raw dataframe
remove(df_cea_raw)
```

3. Inspect the subset 
4. Replace any NAs with 0
```{r echo=TRUE}
# Inspect the cleansing
# Describing the data subset 
head(df_cea_piv)
tail(df_cea_piv)
names(df_cea_piv)
str(df_cea_piv)
summary(df_cea_piv)

# Check the number of Rows & Columns in the data subset
nrow(df_cea_piv) 
ncol(df_cea_piv) 
dim(df_cea_piv) 

# Replace NAs with 0
df_cea_piv[is.na(df_cea_piv)] <- 0
summary(df_cea_piv)
```

## Step 5 - Exploratory Data Analysis

1. Correlation Analysis
```{r echo=TRUE}
# Pairs plot - Correlation plot
df_cea_piv$tract <- as.factor(df_cea_piv$tract)
pairs(df_cea_piv)

rownames(df_cea_piv) <- df_cea_piv$tract

corrmatrix <- cor(df_cea_piv[,2:7])
corrplot(corrmatrix, method = 'number')
```

2. Box plot to check pollutants
```{r echo=TRUE}
# 5.3 - Boxplot - Total Concentration by Pollutant
ggplot(df_cea, aes(x=pollutant, y=total_conc, fill=pollutant)) +
  geom_boxplot()+
  labs(title="Box plot - Total Concentration by Pollutant",x="Pollutant", y = "Concentration")+ scale_fill_brewer(palette="Dark2") + theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

3. Plotting based on observations 
```{r echo=TRUE}
# Toluene vs. 1,3-Butadiene 
plot(Toluene ~ `1,3-Butadiene`, data = df_cea_piv, main = "Toluene vs. 1,3-Butadiene")

# Benzene vs. 1,3-Butadiene 
plot(Benzene ~ `1,3-Butadiene`, data = df_cea_piv, main = "Benzene vs. 1,3-Butadiene")

# Toluene vs. Diesel PM 
plot(Toluene ~ `Diesel Pm`, data = df_cea_piv, main = "Toluene vs. Diesel PM")
```

4. Apply Linear Regression
```{r echo=TRUE}
# Apply the lm() function.
relation <- lm(df_cea_piv$`1,3-Butadiene` ~ df_cea_piv$Benzene)

print(summary(relation))

# Plot the chart.
plot(df_cea_piv$`1,3-Butadiene`, df_cea_piv$Benzene ,col = "blue",  main = "1,3-Butadiene & Benzene - Regression",
abline(lm(df_cea_piv$Benzene ~ df_cea_piv$`1,3-Butadiene`)),cex = 1.3, pch = 16,xlab = "Concentration of Benzene",ylab = "Concentration of 1,3-Butadiene")
```


## Step 6 - Data Preprocessing

1. Create Benchmark variables and classify into High & Low conc. areas
```{r echo=TRUE}
# Create Benchmark Variables to classify the variables
# Anything greater than avg, can be considered as High or else Low conc
toluene_50 <- mean(df_cea_piv$Toluene)
butadiene_50 <- mean(df_cea_piv$`1,3-Butadiene`)
acetaldehyde_50 <- mean(df_cea_piv$Acetaldehyde)
benzene_50 <- mean(df_cea_piv$Benzene)
cyanide_50 <- mean(df_cea_piv$`Cyanide Compounds`)
diesel_50 <- mean(df_cea_piv$`Diesel Pm`)

# Convert continuous values into dichotomous variables (High/Low)
df_cea_piv <- df_cea_piv %>% mutate(Toluene = ifelse(Toluene > toluene_50, "H", "L"))
df_cea_piv <- df_cea_piv %>% mutate(`1,3-Butadiene` = ifelse(`1,3-Butadiene` > butadiene_50, "H", "L"))
df_cea_piv <- df_cea_piv %>% mutate(Acetaldehyde = ifelse(Acetaldehyde > acetaldehyde_50, "H", "L"))
df_cea_piv <- df_cea_piv %>% mutate(Benzene = ifelse(Benzene > benzene_50, "H", "L"))
df_cea_piv <- df_cea_piv %>% mutate(`Cyanide Compounds` = ifelse(`Cyanide Compounds` > cyanide_50, "H", "L"))
df_cea_piv <- df_cea_piv %>% mutate(`Diesel Pm` = ifelse(`Diesel Pm` > diesel_50, "H", "L"))

# Convert to factor
df_cea_piv$Toluene <- as.factor(df_cea_piv$Toluene)
df_cea_piv$`1,3-Butadiene` <- as.factor(df_cea_piv$`1,3-Butadiene`)
df_cea_piv$Acetaldehyde <- as.factor(df_cea_piv$Acetaldehyde)
df_cea_piv$Benzene <- as.factor(df_cea_piv$Benzene)
df_cea_piv$`Cyanide Compounds` <- as.factor(df_cea_piv$`Cyanide Compounds`)
df_cea_piv$`Diesel Pm` <- as.factor(df_cea_piv$`Diesel Pm`)

# Check frequency
as.data.frame(table(df_cea_piv$Toluene)) 
as.data.frame(table(df_cea_piv$`1,3-Butadiene`)) 
as.data.frame(table(df_cea_piv$Acetaldehyde))
as.data.frame(table(df_cea_piv$Benzene)) 
as.data.frame(table(df_cea_piv$`Cyanide Compounds`)) 
as.data.frame(table(df_cea_piv$`Diesel Pm`)) 

# Move these dichotonomous variables into a DF
df_cea_arm <- df_cea_piv[2:7]
skim(df_cea_arm)
```

2. Analyzing the custom-defined concentration levels
```{r echo=TRUE}
# Concentration_Tracts
#colSums() function computes the sums of columns.
high <- colSums(df_cea_arm == "H")
high

low <- colSums(df_cea_arm == "L")
low

df_conc <- rbind(high,low)
df_conc


barplot(df_conc,legend=rownames(df_conc),las=2) #Plot 1

barplot(df_conc, beside=T,legend=rownames(df_conc),las=2)# Plot 2

```

3. Unpivot & Saving the cleaned dataset as a .csv (for SAS)
```{r echo=TRUE}
# 3.1 - Unpivot the Pollutants and risks
df_cea_unpiv <- df_cea_piv %>% 
  pivot_longer(!tract, names_to = "Pollutant", values_to = "Risk Status")

df_cea_unpiv$Potent_Risk <- ifelse(df_cea_unpiv$`Risk Status` == 'L',
                                    "", df_cea_unpiv$Pollutant)

head(df_cea_unpiv)
tail(df_cea_unpiv)

# 3.2 - Write to a .csv file
write.csv(df_cea_unpiv, "CancerRisk_Pollutants.csv", row.names=TRUE)

```


## Step 7 - Association Rules Mining

1. Applying Apriori algorithm
```{r echo=TRUE}
search()

dim(df_cea_arm)
head(df_cea_arm)

all_rules <- apriori(df_cea_arm,
                parameter =list(minlen=2,maxlen=3,conf = 0.70, supp = 0.30))

summary(all_rules)
inspect(all_rules)

rules <- apriori(df_cea_arm,
                parameter =list(minlen=2,maxlen=3,conf = 0.70),
                appearance = list(rhs=c("Acetaldehyde=H"),default="lhs"))

summary(rules)
inspect(rules)

rules <- apriori(df_cea_arm,
                parameter =list(minlen=2,maxlen=3,conf = 0.70),
                appearance = list(rhs=c("Diesel Pm=H"),default="lhs"))

summary(rules)
inspect(rules)

rules <- apriori(df_cea_arm,
                parameter =list(minlen=2,maxlen=3,conf = 0.70),
                appearance = list(rhs=c("1,3-Butadiene=H"),default="lhs"))

summary(rules)
inspect(rules)

rules <- apriori(df_cea_arm,
                parameter =list(minlen=2,maxlen=3,conf = 0.50),
                appearance = list(rhs=c("Cyanide Compounds=H"),default="lhs"))

summary(rules)
inspect(rules)

rules <- apriori(df_cea_arm,
                parameter =list(minlen=2,maxlen=3,conf = 0.70),
                appearance = list(rhs=c("Benzene=H"),default="lhs"))

summary(rules)
inspect(rules)

rules <- apriori(df_cea_arm,
                appearance = list(rhs=c("Toluene=H"),default="lhs"))

summary(rules)
inspect(rules)

rules3 <- apriori(df_cea_arm,
parameter = list(minlen=2,maxlen=3, conf = 0.70),
appearance = list(rhs=c("1,3-Butadiene=H","Diesel Pm=H" ), lhs = c("Acetaldehyde=H","Benzene=H","Toluene=H","Cyanide Compounds=H") , default="none"))

inspect(rules3)

```

2. Using ruleExplorer()
```{r}
ruleExplorer(all_rules)
```



## Step 8 - Model Evaluation
```{r echo=TRUE}
# arulesViz - plots

plot(rules)

plot(rules@quality)
```

